Assesing Health Care Coverage and Access Utilizing the National Health Interview Survey

### 2017-03-08 paper/paper.Rmd

@@

@@

@@

# RELATED WORK
A description of previous papers or projects related to your project.

@@

>>>>

To answer the question, _"Does higher educational attainment cause greater chance of acquiring medical insurance?"_, we filtered the dataset, only looking a three specific variables, the _highest educational attainment_, _health insurance coeverage_, and _health insurance offered at workplace_.

@@

```{r healthcovereduattainment, echo=FALSE}
ggplot(d2015, aes(EDUC1)) + geom_histogram(stat='count') + facet_wrap(~NOTCOV) + 
  ggtitle('Education Level vs. Health Coverage') + labs(x='Education Level') 
```
**Figure 3. Distribution of Health Coverage per Education Level**

```{r workhealthinsuranceeduattainment, echo=FALSE}
ggplot(d2015, aes(EDUC1)) + geom_histogram(stat='count') + facet_wrap(~HIEMPOF) +
  ggtitle('Education Level vs. Workplace-offered Health Insurance') + labs(x='Education Level')
```
**Figure 4. Educational Attainment vs. Workplace-offered Health Insurance**

@@

The first graph that we see here represents whether or not an individual obtained health coverage based on their education level.  Immediately, we see that the number 2 subgraph contains a lot more results than the number 1 subgraph, which indicates that there were more people who received health coverage than did not. Under the number 1 subgraph, we see the varying number of people who did not receive health coverage given their level of education.  There appears to be a notable peak at the number 14 which is indicative of those who have completed up to a high school diploma.  A second notable peak is at the number 15 which represents those who have attended college but had not received a degree.  Such might be the case because most people finish school up until high school and also attend college, so it would make sense that the bulk of the sample surveyed would reside in that educational attainment categories.  It would furthermore make sense that they did not receive health coverage because nowadays, people obtain health/medical insurance from their workplaces since getting these plans on an individual basis is a lot more expensive.  Those who did not finish college or only obtained a high school diploma might have not been able to get the best occupation due to their lack of education and skills in the workforce, which affected their ability to get health coverage from their employer.

In the number 2 subgraph, we see three notable peaks--one at the x-coordinate values 14, 15, and 18.  These peaks correspond to the bulk of those who received health/medical insurance given that they either graduated high school, went to college but did not get a degree, or got a Bachelor’s degree.  What is interesting about this is that there were a lot of people who did not AND did get health coverage given that they graduated from high school or went to college but did not get a degree.  This might reveal to us that even with a high school degree it is possible to get a steady job that includes medical/health insurance coverage.  However, analyzing the data further, it makes sense that a lot of people who received health coverage received at least a Bachelor’s degree, because that level of education makes it a lot easier to get jobs that pay well and include a lot of benefits such as medical insurance.

The second graph we are looking at compares educational attainment with whether or not a person received health insurance from their workplace.  The number 1 subgraph represents the count of people who did _receive health insurance at their workplace_ given their highest level of educational attainment, while number 2 subgraph represents the count of people who _did not receive health insurance at work_ given their highest level of educational attainment.  Again, in both graphs, x-coordinate values 14, 15, and 18, corresponding to having graduated high school, attending college without getting a degree, and getting a bachelor’s degree respectively, are the most prominent peaks in the graph.  This is due to the fact that most people finish at least up until high school, attend college at some point and realize that it wasn’t for them, or obtain their bachelor’s degree.  Very _few_ people quit grade school in the middle of it and very few people get a higher education such as a master’s or doctoral degree, which is why the sample count is so low at those other values.  Based on the number 1 subgraph however, we see that the most people who got health insurance from their workplace are those who graduated with a Bachelor’s degree. This reveals the importance of obtaining a Bachelor’s degree when getting a job.  Other interesting points of analysis are that more high school graduates received workplace-offered health insurance than those who attended college but did not get a degree.  One reasoning behind this is that those who did not receive a degree from college might have been those who quit in order to pursue their own businesses, and therefore did not work for an existing corporation that would have otherwise offered them health insurance.  The number 2 subgraph shows that high school graduates were the least likely to receive health insurance from their workplace, which again points back to the fact that they probably weren’t able to get a great enough job that offered them benefits such as health insurance.  Most decent jobs nowadays require people to have at least a bachelor’s degree.  The lowest of the three peaks, though, are those who got a bachelor’s degree, and it shows that with a higher level of education, you will have a _better_ chance of getting a job with better benefits.

ODDS RATIO: 1.19 / 3.66 = **0.33**
The odds of getting coverage being a high school grad is 0.33 times higher than the odds of getting coverage having a bachelor’s degree.

@@

# RELATED WORK
Previous projects have been done that measured trends in health insurance coverage over the course of time.  The report was done by the National Center for Health Statistics (NCHS), and presents the selected estimates for health insurance coverage among the civilian noninstitutionalized U.S. population based on data from 2015 National Health Interview Survey (NHIS), as well as comparable estimates from the NHIS 2010-2014 surveys.  From their results, they were able to determine that the trends displayed an overall increase in the population that had health insurance coverage in comparison to previous years.  They then observe short term trends by age, poverty status, as well as race and ethnicity to provide a deeper insight of health insurance coverages, and what might be influencing access to them.

@@

3.  Robin A. Cohen. May 2016. Health Insurance Coverage: Early Release of
Estimates From the National Health Interview Survey, 2015. Retrieved March 2, 2017 from https://www.cdc.gov/nchs/data/nhis/earlyrelease/insur201605.pdf
